\section{Can Safety Be Made Corruption-Robust?}
\label{sec:defenses}

\subsection{Safety tuning cracks under corruption}
\label{sec:defenses:tis}

Reasoning-based safety fine-tuning on Think-in-Safety data~\cite{tis2025} applied to LLaVA-CoT cuts SIUO HR\textsubscript{C} from 67.7\% (clean, undefended) to 25.1\% (clean, tuned)\footnote{The undefended clean cell is re-judged independently here and in \cref{tab:siuo-main}, which gives 67.7 and 68.3 --- a one-item difference on $n{=}167$. We quote each run's own value rather than silently harmonizing them; the gap bounds this judge's run-to-run variation on a single cell.} --- and then \emph{every one} of the ten corruptions claws part of it back, by $+1.3$ (motion blur) to $+8.4$ (zoom blur) points (sign test $p{=}0.002$). In relative terms the picture is stark: zoom blur costs the tuned model a third of its safety margin while costing it only 4 points of ScienceQA accuracy. Meanwhile the same tuned model is \emph{immune} to corruption exactly where the harm-location principle predicts: on typographic MM-SafetyBench its ASR is 0\% and on text-borne BeaverTails-style prompts ASR stays flat near 20\% (down from 77\%) under every corruption --- because there the safety trigger never depended on fragile visual evidence. Safety tuning, as currently practiced, hardens the text channel and leaves the image channel soft. It also carries a tax orthogonal to corruption: on benign prompts, over-refusal rises steeply with TIS tuning on both benign suites --- from 59.6\% to 89.7\% on the MMSafeAware~\cite{wang2025mmsafeaware} over-safety subset (428 natural-image items) and from 26.8\% to 74.8\% on the adapted XSTest~\cite{rottger2024xstest} safe subset (250 text-on-image items), where an item counts as over-refused if the model either returns an explicit refusal or judges the benign content unsafe. We report the two separately rather than as one average, for the reason \cref{tab:benign-control} makes precise: the suites differ in where the request lives, so a pooled over-refusal rate mixes a change in compliance with a change in legibility.

\subsection{Where does safety live? A module ablation}
\label{sec:defenses:cat}

\begin{table}[t]
  \centering
  \caption{\textbf{The final safety decision lives in the LLM, not the vision tower.} SIUO under zoom blur (severity 2), GPT-4o judge, $n{=}167$. All four checkpoints are LLaVA-CoT with the same LoRA recipe ($r{=}8$, $\alpha{=}16$); rows differ only in the training \emph{data} (first two) or the \emph{module} the adapter is attached to (last three). Row 1 is the reference safety tune of \cref{sec:defenses:tis}. The corruption-aware rows use a 75\%/25\% zoom-blur(sev 2)/clean copy of the tuning set, so the training corruption matches the evaluation corruption.}
  \label{tab:module}
  \small
  \setlength{\tabcolsep}{4pt}
  \begin{tabular}{llcc}
    \toprule
    Tuning data & LoRA placement & HR\textsubscript{R} & HR\textsubscript{C} \\
    \midrule
    clean (reference)     & LLM only      & 72.5 & 31.1 \\
    \midrule
    zoom blur 75\%        & LLM only      & 67.1 & \textbf{31.7} \\
    zoom blur 75\%        & LLM $+$ vision & 72.5 & 34.7 \\
    zoom blur 75\%        & vision only   & \textbf{50.3} & 50.9 \\
    \bottomrule
  \end{tabular}
\end{table}

Can training the safety tune \emph{on corrupted images} restore robustness? Holding the adapter on the LLM and changing only the data, \cref{tab:module} says: partly, and along one axis only. The reasoning trace becomes meaningfully safer (HR\textsubscript{R} $72.5\rightarrow67.1$) while the final answer does not move (HR\textsubscript{C} $31.1\rightarrow31.7$). A second, independently built recipe --- a uniform four-way mix of clean/zoom/snow/glass images rather than this 75/25 zoom-blur split --- reproduces the same split verdict (HR\textsubscript{R} $-5.6$, HR\textsubscript{C} $+2.4$), so the result is a property of corruption-aware tuning rather than of one mixing ratio. Both pipelines are validated by a clean-only control that reproduces the reference checkpoint's training loss to the fourth decimal (1.5930 vs.\ 1.5929).

The module ablation in the lower block of \cref{tab:module} explains why, and is to our knowledge the first localization of corruption-robust safety within a VLM. Here the data is held fixed and only the adapter's placement changes. Placing the corruption-aware LoRA only on the \emph{vision tower} produces the safest reasoning in the study (HR\textsubscript{R} 50.3 vs.\ 72.5 for the reference) --- adapting perception to corruption genuinely helps the model \emph{recognize} risk in degraded images --- yet its conclusion harmful rate is the \emph{worst} (50.9): the safer perception never reaches the final answer. Conversely, the LLM-only variant carries essentially all of the conclusion-level safety (31.7, matching the reference 31.1). The safety \emph{decision} is implemented in the language model; the vision tower only supplies its trigger. This decomposition sharpens the open problem: corruption-aware training fixes the trigger, but the decision pathway from a risk-aware reasoning trace to a refusal is itself the weakest link --- consistent with the reasoning-leakage finding of \cref{sec:results:reasoning}.

\subsection{Safety prompting: cheap, effective, and only partly what we expected}
\label{sec:defenses:prompt}

\begin{table}[t]
  \centering
  \caption{\textbf{Safety prompting transfers; corruption-awareness does not.} SIUO HR\textsubscript{C} (\%); lower is safer, best prompt per row in bold. \emph{safety} = generic safety nudge; \emph{blur-safe} = the same nudge $+$ an explicit warning that the image may be degraded. Both prompts beat no prompt in all 12 cells, and the generic prompt drives every corrupted condition below that model's own undefended \emph{clean} baseline. But the corruption-aware addition helps only the model it was developed on: it wins 4 of 4 cells on LLaVA-CoT and loses 8 of 8 on the two held-out models.}
  \label{tab:prompting}
  \small
  \begin{tabular}{lccc}
    \toprule
    Condition & no prompt & safety & blur-safe \\
    \midrule
    \multicolumn{4}{l}{\emph{LLaVA-CoT} (prompt developed on this model)} \\
    \quad clean      & 68.3 & 59.9 & \textbf{52.1} \\
    \quad zoom blur  & 74.8 & 62.3 & \textbf{53.9} \\
    \quad snow       & 71.9 & 58.7 & \textbf{50.9} \\
    \quad glass blur & 71.3 & 61.7 & \textbf{56.9} \\
    \midrule
    \multicolumn{4}{l}{\emph{Qwen2.5-VL} (held out)} \\
    \quad clean      & 70.1 & \textbf{49.7} & 56.9 \\
    \quad zoom blur  & 74.2 & \textbf{60.5} & 62.3 \\
    \quad snow       & 73.0 & \textbf{56.3} & 60.5 \\
    \quad glass blur & 73.0 & \textbf{55.7} & 59.3 \\
    \midrule
    \multicolumn{4}{l}{\emph{R1-Onevision} (held out)} \\
    \quad clean      & 83.8 & \textbf{71.9} & 73.0 \\
    \quad zoom blur  & 86.8 & \textbf{71.3} & 81.4 \\
    \quad snow       & 84.4 & \textbf{69.5} & 79.0 \\
    \quad glass blur & 85.6 & \textbf{70.7} & 78.4 \\
    \bottomrule
  \end{tabular}
\end{table}

A system prompt requires no training, no data, and no weight access, so it is the cheapest defense available. We evaluate two --- a generic safety nudge, and the same nudge plus an explicit warning that the image may be degraded --- on three models, having developed both on LLaVA-CoT alone. \Cref{tab:prompting} separates a result that transfers from one that does not.

What transfers is \emph{generic} safety prompting. It lowers HR\textsubscript{C} in all twelve cells, by 8.4 to 20.4 points, and in every model it drives all three corrupted conditions below that model's own undefended clean baseline: a blurred, safety-prompted Qwen2.5-VL (60.5) is safer than a clean, unprompted one (70.1). A deployment that can afford one line of system prompt can buy back more safety than corruption takes away.

What does not transfer is the corruption-aware component --- the part we thought was the contribution. On LLaVA-CoT, warning the model about degradation beat the generic prompt in all four conditions by 4.8--8.4 points. On the two held-out models it \emph{loses} in all eight, by 1.8--7.2 points on Qwen2.5-VL and 1.2--10.2 on R1-Onevision, and the penalty is largest exactly where it was supposed to help most (R1-Onevision under zoom blur, $+10.2$). Twelve cells split 4--8 against the intervention is not a defense; it is a model-specific idiosyncrasy that a single-model evaluation would have published as a general finding.

We offer a mechanism, and flag it as a hypothesis this experiment does not test. Telling a model its image may be unreliable is an invitation to discount the image. On SIUO the harm signal \emph{lives} in the image, so discounting it is precisely the \cref{sec:results:harmloc} failure mode --- resolving degraded evidence as ``unrelated to the image'' --- now induced deliberately by the prompt rather than by the corruption. On LLaVA-CoT that invitation apparently resolves toward caution; on the two reasoning models it appears to resolve toward dismissal. If that reading is right, the corruption-aware framing is not merely inert on unseen models but actively counterproductive on any benchmark whose harm is visual, and the right prompt-side intervention would tell the model to become \emph{more} careful about a degraded image rather than less confident in it. The utility cost of prompting is negligible throughout: on LLaVA-CoT, MathVista moves 53.5$\rightarrow$53.0 and ScienceQA 90.8$\rightarrow$94.0 across the same prompts.

\subsection{Limitations}
\label{sec:defenses:limits}

Our headline metrics flow through a GPT-4o judge; we mitigate with paired statistics, a clean-image Llama-Guard judge for FigStep, and a judge-free refusal-based estimate that recovers the same effect (\cref{sec:setup}), but judge bias cannot be fully excluded. SIUO is small ($n{=}167$); we therefore never interpret single cells, and rest the universality claim on the consistency across models and corruptions (sign test), the dose--response trends, and the independent $n{=}476$ HoliSafe replication (\cref{tab:holisafe}). The corruption-aware fine-tuning result is one recipe (LoRA $r{=}8$, one mixing ratio) on one model, and we expect it to be no more transferable than the prompting result turned out to be (\cref{tab:prompting}) --- a held-out test of it is the most obvious next experiment. The prompting grid covers three models and one benchmark; because the mechanism we propose for the corruption-aware prompt's failure turns on SIUO's harm being visual, that explanation predicts the penalty should shrink or reverse on a rendered-text benchmark, which we have not tested. Finally, all evaluations use synthetic ImageNet-C corruptions; real captured degradations (motion at capture time, sensor noise, compression stacks) are correlated differently and deserve a dedicated benchmark.
